\chapter{Conclusion}

\section{Main results achieved} 
  
  Following \textbf{results} have been achieved:

  \begin{enumerate}

    \item Parser combinator library based on monad transformers that uses special 
    monoids for input stream representation has been developed. Source code of
    library is available on GitHub~\cite{mdParse}.

    \item Prototype of parser combinator library based on extensible effects has
    been developed. Source code of library is available on GitHub~\cite{mdParse}.

    \item Basing on library from point one, parser for subset of Markdown enriched 
    with \LaTeX blocks has been built, together with~\lstinline{HTML} code generator. Source code of implementation and build instructions are also available on 
    Github~\cite{mdParse}.   

  \end{enumerate}

  In addition, section 1.2 contains a comparative analysis of convenience of 
  programming with two approaches to control of computational effects: monad
  transformers and extensible effects.

\section{Possible applications}

  Developed libraries may be used for syntax analysis of markup and programming 
  languages.

  One possible application of Markdown with \LaTeX blocks parser is a electronic 
  lecture notes system. 

\section{Future research}
  
  Extensible effects is a implementation of abstractions of algebraic effects and 
  effects handlers. These abstractions are in its infancy and it could be useful 
  to perform an approbation of its implementations as a machinery for constructing  
  parser combinators libraries.